---
title: "LogFileDataCollection"

author: "Sam Ardery"

date: "`r Sys.Date()`"

output: 
  html_document: 
    toc: true 
    toc_float: true 
    theme: spacelab
    code_folding: hide
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

This document will take a directory containing only the log files of multiple runs of the gbrs pipeline and output the alignment data from the bowtie steps. This document contains hardcoding when the directory is needed so it is important to change these values when running this with a different directory.

# Packages

This code loads all of the packages necessary to run this document
*Note* If you have not already installed these packages, you must do that before running this chunk.
```{r Load-packages , message = FALSE}
library(tidyverse)
library(ggpubr)
```

# Paired-end {.tabset .tabset-pills}

## Extracting the data

Each log file must be opened and the data extracted. When using the gbrs pipeline, the sample id is located on the fourth line and the bowtie alignment data is located on the 7-10 lines, so these were the lines extracted for analysis. At the end of the chunk, the data is written out to a txt file so that this can be saved for later analysis. The txt file will be reloaded so the data vector can be used again for the SE data.

```{r Read-and-isolate-PE-log-data}
dir = "/projects/munger-lab/SampleReruns/PE_Log_Files/"
dataVector <- c()
#Getting the list of files
listOfFiles <- list.files(path = "/projects/munger-lab/SampleReruns/PE_Log_Files/")
#in a loop, reading each file and extracting the sample id and the bowtie read data
for(file in listOfFiles) {
  data <- read.delim(paste0(dir, file))
  samplename <- data[4,]
  if(str_length(samplename) < 20){
      dataVector <- append(dataVector, str_extract(samplename, "PB[:digit:]+_[:digit:]+"))
    } else {
      dataVector <- append(dataVector, str_extract(samplename, "PB[:digit:]+_[:digit:]+_ES_B"))
    }
  dataVector <- append(dataVector, data[7,]) #Processed Reads - the number of reads that bowtie took in to process
  dataVector <- append(dataVector, data[8,]) #Aligned Reads - the number of reads that had at least one alignment
  dataVector <- append(dataVector, data[9,]) #Failed Reads - the number of reads that failed to align
  dataVector <- append(dataVector, data[10,]) #Reported Reads
}
#writing data to a text file
write(dataVector, "PEBowtieData.txt")
```

## Importing and filtering the data

The txt file is read back in and each piece of information is separated from the file and placed into its own vector so that the data can be directly compared.

```{r PE-isolating-required-data}
#importing the data 
df <- tibble(readLines("PEBowtieData.txt"))
#filtering for sample ids
SampleIDs <- filter(df, str_detect(df$`readLines("PEBowtieData.txt")`, "PB"))
SampleIDs <- rename(SampleIDs, SampleIDs = `readLines("PEBowtieData.txt")`)
#filtering for the processed reads
ProcessedReads <- filter(df, str_detect(df$`readLines("PEBowtieData.txt")`, "processed"))
ProcessedReads <- rename(ProcessedReads, ProcessedReads = `readLines("PEBowtieData.txt")`)
#filtering for the aligned reads
AlignedReads <- filter(df, str_detect(df$`readLines("PEBowtieData.txt")`, "least"))
AlignedReads <- rename(AlignedReads, AlignedReads = `readLines("PEBowtieData.txt")`)
#filtering for the failed reads
FailedReads <- filter(df, str_detect(df$`readLines("PEBowtieData.txt")`, "failed"))
FailedReads <- rename(FailedReads, FailedReads = `readLines("PEBowtieData.txt")`)
#filtering for the total reported reads
ReportedReads <- filter(df, str_detect(df$`readLines("PEBowtieData.txt")`, "stream"))
ReportedReads <- rename(ReportedReads, ReportedReads = `readLines("PEBowtieData.txt")`)
```

Sub-setting the data for only the numeric values (this code is just to test, it will be used to directly create the complete data frame)

```{r PE-data-frame-creation-test, results = "hide"}
#extract the number from each read column
as.numeric(str_extract(ProcessedReads$ProcessedReads, "[:digit:]+"))
as.numeric(str_extract(AlignedReads$AlignedReads, "[:digit:]+"))
as.numeric(str_extract(FailedReads$FailedReads, "[:digit:]+"))
as.numeric(str_extract(ReportedReads$ReportedReads, "[:digit:]+"))
```

Creating a completed data frame used the code above

```{r PE-data-frame-creation}
CompletePEdf <- data.frame(SampleID = SampleIDs,
                         Processed = as.numeric(str_extract(ProcessedReads$ProcessedReads, "[:digit:]+")),
                         Aligned = as.numeric(str_extract(AlignedReads$AlignedReads, "[:digit:]+")),
                         Failed = as.numeric(str_extract(FailedReads$FailedReads, "[:digit:]+")),
                         Reported = as.numeric(str_extract(ReportedReads$ReportedReads, "[:digit:]+")))
```

## Analysis

Printing the summary for raw numerical data

```{r PE-data-summary}
#Processed Reads
print("Summary of Processed Reads")
summary(CompletePEdf$Processed)
#Aligned Reads
print("Summary of Aligned Reads")
summary(CompletePEdf$Aligned)
#Failed Reads
print("Summary of Failed Reads")
summary(CompletePEdf$Failed)
#Reported Reads
print("Summary of Reported Reads")
summary(CompletePEdf$Reported)
```

Displaying Aligned and Failed Reads as percentages (as it relates to processed read counts)

```{r PE-adding-data-percentage}
CompletePEdf %>%
  #Get the percentage for aligned reads
  mutate(AlignedPerc = (Aligned / Processed) * 100) %>%
  #Get the percentage for failed reads
  mutate(FailedPerc = (Failed / Processed) * 100) %>%
  #Get the ratio of reported reads to aligned reads
  mutate(ReportRatio = (Reported / Aligned)) %>%
  #Get reported percentage
  mutate(ReportedPerc = (Reported / Processed) * 100) -> CompletePEdf
```

Printing summary of percentages. All percentages are in reference to the quantity of processed reads.

```{r PE-percenage-summary}
#Aligned read percentage
print("Aligned Read percentage")
summary(CompletePEdf$AlignedPerc)
#Failed read percentage
print("Failed Read Percentage")
summary(CompletePEdf$FailedPerc)
#Reported Read Ratio
print("Reported/Aligned Ratio")
summary(CompletePEdf$ReportRatio)
#Reported Read Percentage
print("Reported Read Percentage")
summary(CompletePEdf$ReportedPerc)
```

# Single End Data {.tabset .tabset-pills}

## Extracting the data

Here, the same thing is done as the paired-end data. *Note* Up until the creation of the complete data frame, all of the variables used are the same as the paired-end variables. In this code they are overwritten with the single-end data. If you wish to have this data separated out, the variables used here must be renamed.

```{r Read-and-isolate-SE-data}
dir = "/projects/munger-lab/SampleReruns/SE_Log_Files/"
dataVector <- c()
#Getting the list of files
listOfFile <- list.files(path = "/projects/munger-lab/SampleReruns/SE_Log_Files/")
#in a loop, reading each file and extracting the sample id and the bowtie read data
for(file in listOfFile) {
  data <- read.delim(paste0(dir, file))
  samplename <- data[4,]
  if(str_length(samplename) == 15){
      dataVector <- append(dataVector, str_extract(samplename, "PB[:digit:]+_[:digit:]+"))
    } else {
      dataVector <- append(dataVector, str_extract(samplename, "PB[:digit:]+_[:digit:]+_ES_B"))
    }
  dataVector <- append(dataVector, data[7,]) #Processed Reads - the number of reads that bowtie took in to process
  dataVector <- append(dataVector, data[8,]) #Aligned Reads - the number of reads that had at least one alignment
  dataVector <- append(dataVector, data[9,]) #Failed Reads - the number of reads that failed to align
  dataVector <- append(dataVector, data[10,]) #Reported Reads
}
#writing data to a text file
write(dataVector, "SEBowtieData.txt")
```

## Importing and filtering the data

The txt file is read back in and each piece of information is separated from the file and placed into its own vector so that the data can be directly compared.

```{r SE-isolating-required-data}
#importing the data 
df <- tibble(readLines("SEBowtieData.txt"))
#filtering for sample ids
SampleIDs <- filter(df, str_detect(df$`readLines("SEBowtieData.txt")`, "PB"))
SampleIDs <- rename(SampleIDs, SampleIDs = `readLines("SEBowtieData.txt")`)
#filtering for the processed reads
ProcessedReads <- filter(df, str_detect(df$`readLines("SEBowtieData.txt")`, "processed"))
ProcessedReads <- rename(ProcessedReads, ProcessedReads = `readLines("SEBowtieData.txt")`)
#filtering for the aligned reads
AlignedReads <- filter(df, str_detect(df$`readLines("SEBowtieData.txt")`, "least"))
AlignedReads <- rename(AlignedReads, AlignedReads = `readLines("SEBowtieData.txt")`)
#filtering for the failed reads
FailedReads <- filter(df, str_detect(df$`readLines("SEBowtieData.txt")`, "failed"))
FailedReads <- rename(FailedReads, FailedReads = `readLines("SEBowtieData.txt")`)
#filtering for the total reported reads
ReportedReads <- filter(df, str_detect(df$`readLines("SEBowtieData.txt")`, "stream"))
ReportedReads <- rename(ReportedReads, ReportedReads = `readLines("SEBowtieData.txt")`)
```

Sub-setting the data for only the numeric values (this code is just to test, it will be used to directly create the complete data frame)

```{r SE-data-frame-creation-test, results = "hide"}
#extract the number from each read column
as.numeric(str_extract(ProcessedReads$ProcessedReads, "[:digit:]+"))
as.numeric(str_extract(AlignedReads$AlignedReads, "[:digit:]+"))
as.numeric(str_extract(FailedReads$FailedReads, "[:digit:]+"))
as.numeric(str_extract(ReportedReads$ReportedReads, "[:digit:]+"))
```

Creating a completed data frame used the code above

```{r SE-data-frame-creation}
CompleteSEdf <- data.frame(SampleID = SampleIDs,
                         Processed = as.numeric(str_extract(ProcessedReads$ProcessedReads, "[:digit:]+")),
                         Aligned = as.numeric(str_extract(AlignedReads$AlignedReads, "[:digit:]+")),
                         Failed = as.numeric(str_extract(FailedReads$FailedReads, "[:digit:]+")),
                         Reported = as.numeric(str_extract(ReportedReads$ReportedReads, "[:digit:]+")))
```

## Analysis

Printing the summary for raw numerical data

```{r SE-data-summary}
#Processed Reads
print("Summary of Processed Reads")
summary(CompleteSEdf$Processed)
#Aligned Reads
print("Summary of Aligned Reads")
summary(CompleteSEdf$Aligned)
#Failed Reads
print("Summary of Failed Reads")
summary(CompleteSEdf$Failed)
#Reported Reads
print("Summary of Reported Reads")
summary(CompleteSEdf$Reported)
```

Displaying Aligned and Failed Reads as percentages (as it relates to processed read counts)

```{r SE-adding-percentage-data}
CompleteSEdf %>%
  #Get the percentage for aligned reads
  mutate(AlignedPerc = (Aligned / Processed) * 100) %>%
  #Get the percentage for failed reads
  mutate(FailedPerc = (Failed / Processed) * 100) %>%
  #Get the ratio of reported reads to aligned reads
  mutate(ReportRatio = (Reported / Aligned)) %>%
  #Get reported percentage
  mutate(ReportedPerc = (Reported / Processed) * 100) -> CompleteSEdf
```

Printing summary of percentages. All percentages are in reference to the quantity of processed reads.

```{r SE-percenage-summary}
#Aligned read percentage
print("Aligned Read percentage")
summary(CompleteSEdf$AlignedPerc)
#Failed read percentage
print("Failed Read Percentage")
summary(CompleteSEdf$FailedPerc)
#Reported Read Ratio
print("Reported/Aligned Ratio")
summary(CompleteSEdf$ReportRatio)
#Reported Read Percentage
print("Reported Read Percentage")
summary(CompleteSEdf$ReportedPerc)
```

# STAR Single End Data {.tabset .tabset-pills}

## Extracting the data

Here we do a very similar extraction as previously but with slight alterations due to the differences in how Bowtie and STAR report their log data. *Note* Once again, some of the variables used here are the same as used above. If you would like to isolate these variables from the previous two sets, you have to first rename them.
```{r Read-and-isolate-star-data}
dir = "/projects/munger-lab/SampleReruns/STAR/STARLogFiles/"
dataVector <- c()
listOfFiles <- list.files(path = "/projects/munger-lab/SampleReruns/STAR/STARLogFiles/")

for(file in listOfFiles) {
  data <- read.delim(paste0(dir, file))
  if(str_length(file) == 25){
    samplename <- str_extract(file, "PB[:digit:]+_[:digit:]+")
  } else {
    samplename <- str_extract(file, "PB[:digit:]+_[:digit:]+_ES_B")
  }

  dataVector <- append(dataVector, samplename)
  dataVector <- append(dataVector, data[8,2])

}
write(dataVector, "SESTARData.txt")
```


## Importing and filtering the data

Here the data is loaded in from the text file made in the previous chunk and it is filtered down to a data frame that contains one column of all the sample IDs and one column of the alignment percentages. 
```{r Isolating-required-data}
df <- tibble(readLines("SESTARData.txt"))

SampleIDs <- filter(df, str_detect(df$`readLines("SESTARData.txt")`, "P"))
SampleIDs <- rename(SampleIDs, SampleID = `readLines("SESTARData.txt")`)
SampleIDs[120,] <- "PB361_90_ES_B"

AlignedPerc <- filter(df, str_detect(df$`readLines("SESTARData.txt")`, "^[:digit:]"))
AlignedPerc$`readLines("SESTARData.txt")` <- str_extract(AlignedPerc$`readLines("SESTARData.txt")`, "[:digit:][:digit:]\\.[:digit:][:digit:]")
AlignedPerc <- rename(AlignedPerc, SE_STAR = `readLines("SESTARData.txt")`)

STARdf <- data.frame(SampleID = SampleIDs, SE_STAR = AlignedPerc)
STARdf$SE_STAR <- as.numeric(STARdf$SE_STAR) 
```

## Analysis

Printing the summary for the STAR alignment data
```{r STAR-data-summary}
#Aligned Reads
print("Summary of Aligned Reads")
summary(STARdf$SE_STAR)
```


# Combined Analysis 

In this chunk, the alignment percentage is isolated from each data frame and joined together into one data frame that contains the alignment for all three methods for each of the 185 samples. 
```{r Full-alignment-percentage-dataframe}
PEAligned <- data.frame(SampleID = CompletePEdf$SampleID, PE_Bowtie = CompletePEdf$AlignedPerc)
SEAligned <- data.frame(SampleID = CompleteSEdf$SampleID, SE_Bowtie = CompleteSEdf$AlignedPerc)

AlignedPercentages <- full_join(PEAligned, SEAligned, by = "SampleID")
AlignedPercentages <- full_join(AlignedPercentages, STARdf, by = "SampleID")

write_csv(AlignedPercentages, "AlignedPercentages.csv")
```

We can then use this data frame to plot the three alignment methods against each other, showing how well each method aligns reads in comparison to the other methods. 
```{r Alignment-percentage-graphic}
AlignedPercentages %>%
  pivot_longer(c(2:4), names_to = "Alignment_Method", values_to = "Percent_Aligned") %>%
  ggplot(aes(x = Alignment_Method, y = Percent_Aligned)) +
    geom_violin() +
    geom_boxplot(width = 0.1) +
    theme_pubr(base_size = 18) +
    stat_compare_means(label.y = 95) +
    stat_compare_means(label = "p.signif", method = "t.test", ref.group = ".all.") +
    labs(x = "Alignment Method", y = "Percent Aligned")
```

We can then isolate the average of each row and print as a table to show the difference in the average alignment percentage between each data frame. 
```{r Alignment-averages}
Averages <- data.frame(PEBowtie = mean(AlignedPercentages$PE_Bowtie), 
                       SEBowtie = mean(AlignedPercentages$SE_Bowtie), 
                       SESTAR = mean(AlignedPercentages$SE_STAR))

print.data.frame(Averages)
```












